---
title: "Exercise5_group5"
author: "Chelsea"
date: "2/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
##### setup -----
rm(list = ls())
gc()

library(tidyverse)
library(skimr)
library(lubridate)
library(arrow)
library(modelsummary)
library(gender)
set.seed(10101)

##### paths -----
data_path = "/Users/cheukyeehon/Documents/McGill MMA/04 Winter I 2022/ORGB 671 Talent Analytics/USPTO_data/"
```

## Transitions by seniority and gender

We start with `applications` data file, because it's the only source of data on turnover. We load the file and add a gender column based on the examiner's name.

```{r set up data}
## load applications data
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet")) %>% 
  rename(patex_id = examiner_id)

## get gender 
examiner_gender <- applications %>% 
  mutate(
    ff_name = str_extract(examiner_name_first,"\\w+"), # extract first first name
  ) %>% 
  distinct(ff_name) %>% 
  do(results = gender(.$ff_name, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    ff_name = name,
    gender,
    proportion_female
  ) %>% 
  mutate(woman = if_else(gender=="female",1,0))

## add gender back to the applications data
applications <- applications %>% 
  mutate(
    ff_name = str_extract(examiner_name_first,"\\w+"), # extract first first name
  ) %>% 
  left_join(examiner_gender, by = "ff_name")

## How many are missing?
applications %>% select(woman) %>% skim()
```

### Add examiner seniority

Next, we can infer examiner tenure based on first/last observed date of application. This will help us figure out junior/senior switch as well as `onboard` and `exit` dates (if the examiner leaves).

```{r add tenure, skimr_include_summary = FALSE}
## use timestamps on applications to approximate employment dates
examiner_tenure <- applications %>% 
  mutate(    
    filing_date_p = ymd(filing_date), # need to standardize data representation
    status_date_p = dmy_hms(appl_status_date) # so that min()/max() work properly
  ) %>% 
  group_by(patex_id) %>%
  summarise(
    min_filing_date = min(filing_date_p, na.rm = TRUE),
    max_filing_date = max(filing_date_p, na.rm = TRUE),
    min_status_date = min(status_date_p, na.rm = TRUE),
    max_status_date = max(status_date_p, na.rm = TRUE),
    ) %>% 
  rowwise() %>% # find minimum/maximum by row
  mutate(
    onboard_date = min(c(min_filing_date, min_status_date), na.rm = TRUE),
    exit_date = max(c(max_filing_date, max_status_date), na.rm = TRUE)
  ) %>% # clean up some date typos
  mutate(
    onboard_date = if_else(year(onboard_date)>2017 | year(onboard_date)<2000, NA_Date_, onboard_date),
    exit_date =    if_else(year(exit_date)>2017    | year(exit_date)<2000, NA_Date_,    exit_date   ),
    tenure = as.duration(onboard_date %--% exit_date) / dyears(1)
  ) %>% 
  select(
    patex_id,
    onboard_date,
    exit_date,
    tenure
  ) %>% 
  ungroup()

## plot tenure
examiner_tenure %>% 
  select(onboard_date,exit_date,tenure) %>% 
  skim()
```
Let's find those examiners who left the agency. Of course, we have both left and right truncation (censorship) in our sample, so if someone's last observed date is toward the end of our sample's time window, that person may still be employed when our observation stops. So let's mark those whom we don't observe for at least 6 months as "left" the organization. For them, we'll keep the exit date. For the rest, will change the exit date to missing, so that we don't confuse ourselves.

```{r identify exits, skimr_include_summary = FALSE}
## get min and max dates in the data to establish our sample time window
min_date <- examiner_tenure %>% 
  ungroup() %>% 
  summarise(m = min(onboard_date, na.rm = TRUE)) %>% 
  pull()

max_date <- examiner_tenure %>% 
  ungroup() %>% 
  summarise(m = max(onboard_date, na.rm = TRUE)) %>% 
  pull()

## only keep exit_date for those whom we don't see working for at least 6 months
examiner_tenure <- examiner_tenure %>% 
  mutate(
    exit_date = if_else(as.duration(exit_date %--% max_date) / dyears(1) > 0.5,
                        exit_date,
                        NA_Date_)
  )

examiner_tenure %>% 
  select(onboard_date,exit_date,tenure) %>% 
  skim()
```
### Calculate the transitions

Let's assume a time period for our transition matrix of $t_0 = 2007$ (the year we first get senior examiners, according to our definition) and $t_1 = 2011$ (about a mid-point in our time window). For each of the years, we need a vector of indicators:
  + examiner is junior
  + examiner is senior
  + examiner is a new hire <span style="color:blue">during this period </span>
  + examiner exited <span style="color:blue">during this period </span>
  
<span style="color:blue">Note that examiners who have already exit in previous periods will have state NA subsequently.</span>
  
```{r transitions 1}
   transitions <- examiner_tenure %>% 
  mutate(
    tenure_t0 = as.duration(onboard_date %--% ymd("2007-01-01")) / dyears(1),
    tenure_t1 = as.duration(onboard_date %--% ymd("2011-01-01")) / dyears(1),
    tenure_t2 = as.duration(onboard_date %--% ymd("2015-01-01")) / dyears(1),
    t0_state = case_when(
      tenure_t0<6 & tenure_t0>0 ~ "Junior"  , # Jr; not those yet to be hired!
      tenure_t0>=6              ~ "Senior"  , # Sr
      TRUE                      ~ NA_character_ # not yet hired
    ),
    t1_state = case_when(
      exit_date<ymd("2011-01-01")        ~ "Exit",
      onboard_date>ymd("2007-01-01") 
        & onboard_date<ymd("2011-01-01") ~ "New hire",
      tenure_t1<6 & tenure_t1>0          ~ "Junior"  , # Jr; not those yet to be hired!
      tenure_t1>=6                       ~ "Senior"  , # Sr
      TRUE                               ~ NA_character_ # not yet hired
      ),
    t2_state = case_when(
      t1_state=="Exit"                   ~ NA_character_,
      exit_date<ymd("2015-01-01")        ~ "Exit",
      onboard_date>ymd("2011-01-01") 
        & onboard_date<ymd("2015-01-01") ~ "New hire",
      tenure_t2<6 & tenure_t2>0          ~ "Junior"  , # Jr; not those yet to be hired!
      tenure_t2>=6                       ~ "Senior"  , # Sr
      TRUE                               ~ NA_character_ # not yet hired or already exit in previous period
      )
    )
```

<span style="color:blue">Now that we've identified an individual state at $t_0$, $t_1$ and $t_2$, we can spread dates into columns and keep each row per examiner corresponding to a single run of Markov chain.</span>

```{r}
transitions_seniority <- as.data.frame(select(transitions,c("patex_id","t0_state","t1_state","t2_state"))) %>% 
  drop_na(patex_id)
names(transitions_seniority) <- c("patex_id","t0","t1","t2")
rownames(transitions_seniority) <- transitions_seniority$patex_id
transitions_seniority <- select(transitions_seniority,-c("patex_id"))
head(transitions_seniority,10)
```


<span style="color:blue">We can use the following function to compute the overall seniority transition matrix by calculating the transition combinations from one state to another as share of previous period. The horizontal columns refer to the status before and the vertical rownames refer to the status after </span>

```{r}
# Function to compute first-order Markov transition matrix 
trans.matrix <- function(X)
{
  tt <- table( c(X[,-ncol(X)]), c(X[,-1]) )
  tt <- tt / rowSums(tt)
  t(tt)
}

# Function to reorder matrix to logical employment journey
trans.reorder <- function(X){
  order <- c("Junior", "Senior", "Exit")
  X <- X %>%
  slice(match(order, rownames(X)))
  #X <- X[,order]
  X <- X %>% mutate(Exit=c(0,0,1)) # to hard code given only 3 periods are considered
  X <- X %>% mutate(across(where(is.numeric), ~ round(., 4)))  
}

# Create transition matrix (for t0 to t1, all)
trans_matrix_table <- trans.matrix(as.matrix(transitions_seniority[,c(1,2)]))
trans_matrix_all <- as.data.frame.matrix(trans_matrix_table) 
trans_matrix_all <- trans.reorder(trans_matrix_all)
trans_matrix_all

rm(transitions_seniority)

```

<span style="color:blue">Repeat the matrix generation for women.</span>
```{r}
# First to look up gender from applications
id_gender <- unique(select(applications,c("patex_id","woman")))
transitions <- left_join(transitions, id_gender, by="patex_id", match="first")
rm(id_gender)

# Similarly, spread dates into columns and keep each row per examiner corresponding to a single run of Markov chain
transitions_seniority_women <- filter(transitions, woman==1)
transitions_seniority_women <- as.data.frame(select(transitions_seniority_women,c("patex_id","t0_state","t1_state","t2_state"))) %>% 
  drop_na(patex_id)
names(transitions_seniority_women) <- c("patex_id","t0","t1","t2")
rownames(transitions_seniority_women) <- transitions_seniority_women$patex_id
transitions_seniority_women <- select(transitions_seniority_women,-c("patex_id"))
head(transitions_seniority_women,10)

# Create transition matrix (for t0 to t1, women)
trans_matrix_table <- trans.matrix(as.matrix(transitions_seniority_women[,c(1,2)]))
trans_matrix_women <- as.data.frame.matrix(trans_matrix_table) 
trans_matrix_women <- trans.reorder(trans_matrix_women)
trans_matrix_women

rm(transitions_seniority_women)
```

<span style="color:blue">Repeat the matrix generation for men this time.</span>
```{r}
# Similarly, spread dates into columns and keep each row per examiner corresponding to a single run of Markov chain
transitions_seniority_men <- filter(transitions, woman==0)
transitions_seniority_men <- as.data.frame(select(transitions_seniority_men,c("patex_id","t0_state","t1_state","t2_state"))) %>% 
  drop_na(patex_id)
names(transitions_seniority_men) <- c("patex_id","t0","t1","t2")
rownames(transitions_seniority_men) <- transitions_seniority_men$patex_id
transitions_seniority_men <- select(transitions_seniority_men,-c("patex_id"))
head(transitions_seniority_men,10)

# Create transition matrix (for t0 to t1, women)
trans_matrix_table <- trans.matrix(as.matrix(transitions_seniority_men[,c(1,2)]))
trans_matrix_men <- as.data.frame.matrix(trans_matrix_table) 
trans_matrix_men <- trans.reorder(trans_matrix_men)
trans_matrix_men

rm(transitions_seniority_men)
```




<span style="color:red">========================</span>


Now that we've identified an individual's state at $t_0$, $t_1$ and $t_2$, we can calculate transitions from one state to another as shares of the two initial categories (Junior and Senior at $t_0$).

Note that some examiners have negative tenure at our chosen time points, which means they have not been hired yet. We should be careful with how to count them!

```{r transitions 2}
transitions %>% 
  count(t0_state, t1_state, name = "cell") %>%  # get counts by permutation
  group_by(t0_state) %>% # make sure the sum() statement below refers to each t_0 state
  mutate(
    rate = cell/sum(cell) # share of t_0 category total that's in the current t_1 state
    ) %>% 
  ungroup()
```

As you can see, we've accounted for all possible permutations. Or did we? What's missing here?

We can now create the transition matrix:

```{r transitions 3}
tr_matrix <- transitions %>% 
  count(t0_state, t1_state, name = "cell") %>% 
  group_by(t0_state) %>% 
  mutate(
    rate = cell/sum(cell)
    ) %>% 
  ungroup() %>% 
  filter(!is.na(t0_state), t1_state %in% c("Junior","Senior")) %>% 
  select(
    from_state = t0_state, 
    to_state = t1_state, 
    rate
    )

tr_matrix
```

## Predition for $t_2$

Let's count actual cells for $t_2$. 

```{r t2 states, warning=FALSE}
transitions <- transitions %>% 
  mutate(
    tenure_t2 = as.duration(onboard_date %--% ymd("2015-01-01")) / dyears(1),
    t2_state = case_when(
      exit_date<ymd("2015-01-01") &
       exit_date>ymd("2011-01-01")       ~ "Exit",
      onboard_date>=ymd("2011-01-01") 
        & onboard_date<ymd("2015-01-01") ~ "New hire",
      tenure_t2<6 & tenure_t2>0          ~ "Junior"  , # Jr; not those yet to be hired!
      tenure_t2>=6                       ~ "Senior"  , # Sr
      TRUE                               ~ NA_character_ # not yet hired
      )
    )

t1_counts <- transitions %>% 
  count(t1_state, name = "t1_count") %>% 
  filter(t1_state %in% c("Junior","Senior")) %>% 
  rename(from_state = t1_state) %>% 
  left_join(tr_matrix) %>% 
  relocate(from_state, to_state) %>%  # move the column to be the first one, for readability
  mutate(pred_t2 = round(t1_count*rate))

t1_counts
```

Compare with actual:

```{r}
t1_counts %>% 
  group_by(to_state) %>% 
  tally(pred_t2)

transitions %>% 
  count(t2_state)
```


Another way to compare is to calculate a transition matrix for $t_1 \rightarrow t_2$ change.

```{r}
transitions %>% 
  count(t1_state, t2_state, name = "cell") %>%  # get counts by permutation
  group_by(t1_state) %>% # make sure the sum() statement below refers to each t_1 state
  mutate(
    rate = cell/sum(cell) # share of t_0 category total that's in the current t_2 state
    ) %>% 
  ungroup()
```

