---
title: "Exercise5_group5"
author: "Chelsea, Matthew"
date: "2/14/2022"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
##### setup -----
rm(list = ls())
gc()

library(tidyverse)
library(skimr)
library(lubridate)
library(arrow)
#library(modelsummary_all)
library(gender)
set.seed(10101)

##### paths -----
data_path = "/Users/cheukyeehon/Documents/McGill MMA/04 Winter I 2022/ORGB 671 Talent Analytics/USPTO_data/"
```

## Transitions by seniority and gender

We start with `applications` data file, because it's the only source of data on turnover. We load the file and add a gender column based on the examiner's name.

```{r set up data}
## load applications data
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet")) %>% 
  rename(patex_id = examiner_id)

## get gender 
examiner_gender <- applications %>% 
  mutate(
    ff_name = str_extract(examiner_name_first,"\\w+"), # extract first first name
  ) %>% 
  distinct(ff_name) %>% 
  do(results = gender(.$ff_name, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    ff_name = name,
    gender,
    proportion_female
  ) %>% 
  mutate(woman = if_else(gender=="female",1,0))

## add gender back to the applications data
applications <- applications %>% 
  mutate(
    ff_name = str_extract(examiner_name_first,"\\w+"), # extract first first name
  ) %>% 
  left_join(examiner_gender, by = "ff_name")

## How many are missing?
applications %>% select(woman) %>% skim()
```

### Add examiner seniority

Next, we can infer examiner tenure based on first/last observed date of application. This will help us figure out junior/senior switch as well as `onboard` and `exit` dates (if the examiner leaves).

```{r add tenure, skimr_include_summary_all = FALSE}
## use timestamps on applications to approximate employment dates
examiner_tenure <- applications %>% 
  mutate(    
    filing_date_p = ymd(filing_date), # need to standardize data representation
    status_date_p = dmy_hms(appl_status_date) # so that min()/max() work properly
  ) %>% 
  group_by(patex_id) %>%
  summarise(
    min_filing_date = min(filing_date_p, na.rm = TRUE),
    max_filing_date = max(filing_date_p, na.rm = TRUE),
    min_status_date = min(status_date_p, na.rm = TRUE),
    max_status_date = max(status_date_p, na.rm = TRUE),
    ) %>% 
  rowwise() %>% # find minimum/maximum by row
  mutate(
    onboard_date = min(c(min_filing_date, min_status_date), na.rm = TRUE),
    exit_date = max(c(max_filing_date, max_status_date), na.rm = TRUE)
  ) %>% # clean up some date typos
  mutate(
    onboard_date = if_else(year(onboard_date)>2017 | year(onboard_date)<2000, NA_Date_, onboard_date),
    exit_date =    if_else(year(exit_date)>2017    | year(exit_date)<2000, NA_Date_,    exit_date   ),
    tenure = as.duration(onboard_date %--% exit_date) / dyears(1)
  ) %>% 
  select(
    patex_id,
    onboard_date,
    exit_date,
    tenure
  ) %>% 
  ungroup()

## plot tenure
examiner_tenure %>% 
  select(onboard_date,exit_date,tenure) %>% 
  skim()
```
Let's find those examiners who left the agency. Of course, we have both left and right truncation (censorship) in our sample, so if someone's last observed date is toward the end of our sample's time window, that person may still be employed when our observation stops. So let's mark those whom we don't observe for at least 6 months as "left" the organization. For them, we'll keep the exit date. For the rest, will change the exit date to missing, so that we don't confuse ourselves.

```{r identify exits, skimr_include_summary_all = FALSE}
## get min and max dates in the data to establish our sample time window
min_date <- examiner_tenure %>% 
  ungroup() %>% 
  summarise(m = min(onboard_date, na.rm = TRUE)) %>% 
  pull()

max_date <- examiner_tenure %>% 
  ungroup() %>% 
  summarise(m = max(onboard_date, na.rm = TRUE)) %>% 
  pull()

## only keep exit_date for those whom we don't see working for at least 6 months
examiner_tenure <- examiner_tenure %>% 
  mutate(
    exit_date = if_else(as.duration(exit_date %--% max_date) / dyears(1) > 0.5,
                        exit_date,
                        NA_Date_)
  )

examiner_tenure %>% 
  select(onboard_date,exit_date,tenure) %>% 
  skim()
```

### Calculate the transitions

Let's assume a time period for our transition matrix of $t_0 = 2007$ (the year we first get senior examiners, according to our definition), $t_1 = 2011$ (about a mid-point in our time window) for the projection of $t_2 = 2015$. For each of the years, we need a vector of indicators:
  + examiner is junior
  + examiner is senior
  + examiner is a new hire <span style="color:blue">during this period </span>
  + examiner exited <span style="color:blue">during this period </span>
  
<span style="color:blue">Note that examiners who have already exit in previous periods will have state NA subsequently.</span>
  
```{r transitions 1}
   transitions <- examiner_tenure %>% 
  mutate(
    tenure_t0 = as.duration(onboard_date %--% ymd("2007-01-01")) / dyears(1),
    tenure_t1 = as.duration(onboard_date %--% ymd("2011-01-01")) / dyears(1),
    tenure_t2 = as.duration(onboard_date %--% ymd("2015-01-01")) / dyears(1),
    t0_state = case_when(
      tenure_t0<6 & tenure_t0>0 ~ "Junior"  , # Jr; not those yet to be hired!
      tenure_t0>=6              ~ "Senior"  , # Sr
      TRUE                      ~ NA_character_ # not yet hired
    ),
    t1_state = case_when(
      exit_date<ymd("2011-01-01")        ~ "Exit",
      onboard_date>ymd("2007-01-01") 
        & onboard_date<ymd("2011-01-01") ~ "New hire",
      tenure_t1<6 & tenure_t1>0          ~ "Junior"  , # Jr; not those yet to be hired!
      tenure_t1>=6                       ~ "Senior"  , # Sr
      TRUE                               ~ NA_character_ # not yet hired
      ),
    t2_state = case_when(
      t1_state=="Exit"                   ~ NA_character_,
      exit_date<ymd("2015-01-01")        ~ "Exit",
      onboard_date>ymd("2011-01-01") 
        & onboard_date<ymd("2015-01-01") ~ "New hire",
      tenure_t2<6 & tenure_t2>0          ~ "Junior"  , # Jr; not those yet to be hired!
      tenure_t2>=6                       ~ "Senior"  , # Sr
      TRUE                               ~ NA_character_ # not yet hired or already exit in previous period
      )
    )
```

<span style="color:blue">Now that we've identified an individual state at $t_0$, $t_1$ and $t_2$, we can spread dates into columns and keep each row per examiner corresponding to a single run of Markov chain.</span>

```{r}
transitions_seniority <- as.data.frame(select(transitions,c("patex_id","t0_state","t1_state","t2_state"))) %>% 
  drop_na(patex_id)
names(transitions_seniority) <- c("patex_id","t0","t1","t2")
rownames(transitions_seniority) <- transitions_seniority$patex_id
transitions_seniority <- select(transitions_seniority,-c("patex_id"))
head(transitions_seniority,10)
```

### Transitions matrix - general
<span style="color:blue">We can use the following function to compute the overall seniority transition matrix by calculating the transition combinations from one state to another as share of previous period. The horizontal columns refer to the status before and the vertical rownames refer to the status after </span>

```{r}
# Function to compute first-order Markov transition matrix 
trans.matrix <- function(X)
{
  tt <- table( c(X[,-ncol(X)]), c(X[,-1]) )
  tt <- tt / rowSums(tt)
  t(tt)
}

# Function to reorder matrix to logical employment journey
order <- c("Junior", "Senior", "Exit")
trans.reorder <- function(X){
  X <- X %>%
  slice(match(order, rownames(X)))
  #X <- X[,order]
  X <- X %>% mutate(Exit=c(0,0,1)) # to hard code given only 3 periods are considered
  X <- X %>% mutate(across(where(is.numeric), ~ round(., 4)))  
}

# Create transition matrix (for t0 to t1, all)
trans_matrix_table <- trans.matrix(as.matrix(transitions_seniority[,c(1,2)]))
trans_matrix_all <- as.data.frame.matrix(trans_matrix_table) 
trans_matrix_all <- trans.reorder(trans_matrix_all)
trans_matrix_all

rm(transitions_seniority)

```

Comparing junior and senior, it is observed that junior has a higher exit rate of 11.9% than senior 6.3%. 
Next, we will further drill into the impact of gender on promotion and retention. 

### Transitions matrix - women
<span style="color:blue">Repeat the matrix generation for women.</span>
```{r}
# First to look up gender from applications
id_gender <- unique(select(applications,c("patex_id","woman")))
transitions <- left_join(transitions, id_gender, by="patex_id", match="first")
rm(id_gender)

# Similarly, spread dates into columns and keep each row per examiner corresponding to a single run of Markov chain
transitions_seniority_women <- filter(transitions, woman==1)
transitions_seniority_women <- as.data.frame(select(transitions_seniority_women,c("patex_id","t0_state","t1_state","t2_state"))) %>% 
  drop_na(patex_id)
names(transitions_seniority_women) <- c("patex_id","t0","t1","t2")
rownames(transitions_seniority_women) <- transitions_seniority_women$patex_id
transitions_seniority_women <- select(transitions_seniority_women,-c("patex_id"))
#head(transitions_seniority_women,10)

# Create transition matrix (for t0 to t1, women)
trans_matrix_table <- trans.matrix(as.matrix(transitions_seniority_women[,c(1,2)]))
trans_matrix_women <- as.data.frame.matrix(trans_matrix_table) 
trans_matrix_women <- trans.reorder(trans_matrix_women)
trans_matrix_women

rm(transitions_seniority_women)
```
### Transitions matrix - men
<span style="color:blue">Repeat the matrix generation for men this time.</span>
```{r}
# Similarly, spread dates into columns and keep each row per examiner corresponding to a single run of Markov chain
transitions_seniority_men <- filter(transitions, woman==0)
transitions_seniority_men <- as.data.frame(select(transitions_seniority_men,c("patex_id","t0_state","t1_state","t2_state"))) %>% 
  drop_na(patex_id)
names(transitions_seniority_men) <- c("patex_id","t0","t1","t2")
rownames(transitions_seniority_men) <- transitions_seniority_men$patex_id
transitions_seniority_men <- select(transitions_seniority_men,-c("patex_id"))
#head(transitions_seniority_men,10)

# Create transition matrix (for t0 to t1, women)
trans_matrix_table <- trans.matrix(as.matrix(transitions_seniority_men[,c(1,2)]))
trans_matrix_men <- as.data.frame.matrix(trans_matrix_table) 
trans_matrix_men <- trans.reorder(trans_matrix_men)
trans_matrix_men

rm(transitions_seniority_men)
```


Comparing women and men, it is observed that promotion rate (from junior to senior) is higher in women (63.9%) than in men (61.9%) and attrition rate (both from junior to exit and senior to exit) is lower in women (11.4%; 6.3%) than in men (12.4%; 6.5%). 

## Prediction for $t_2$ - general
<span style="color:blue">Now that all 3 matrices are ready. Let's count actual cells for $t_1$.</span>
```{r}
t1_counts <- transitions %>% 
  count(t1_state, name = "t1_count") %>% 
  filter(t1_state %in% c("Junior","Senior","Exit")) %>% 
  rename(from_state = t1_state) %>% 
  slice(match(order, from_state))

t1_counts <- as.data.frame(t1_counts)
rownames(t1_counts) <- t1_counts$from_state
t1_counts <- select(t1_counts,-c("from_state"))

summary_all <- t1_counts
summary_all

rm(t1_counts)
```

<span style="color:blue">Make prediction for $t_2$.</span>
```{r}
mmult <- as.matrix(trans_matrix_all) %*% as.matrix(summary_all)

summary_all <- summary_all %>%
  mutate(
    t2_pred = unlist(lapply(seq_len(ncol(mmult)), function(i) mmult[,i]))
  ) 
summary_all
```

<span style="color:blue">Count the actual cells for $t_2$.</span>
```{r}
t2_counts <- transitions %>% 
  count(t2_state, name = "t2_count") %>% 
  filter(t2_state %in% c("Junior","Senior","Exit")) %>% 
  rename(from_state = t2_state) %>% 
  slice(match(order, from_state))

t2_counts <- as.data.frame(t2_counts)
rownames(t2_counts) <- t2_counts$from_state
t2_counts <- select(t2_counts,-c("from_state"))
t2_counts

summary_all <- cbind(summary_all, t2_counts)
summary_all

rm(t2_counts)
```
<span style="color:blue">Calculate the % difference.</span>
```{r}
summary_all <- summary_all %>%
  mutate(
    percent_diff = (t2_pred - t2_count)/t2_count
  ) 
summary_all
```


## Prediction for $t_2$ - women
<span style="color:blue">Zooming into women, let's count actual cells for $t_1$.</span>
```{r}
t1_counts <- transitions %>% 
  filter(woman==1) %>%
  count(t1_state, name = "t1_count") %>% 
  filter(t1_state %in% c("Junior","Senior","Exit")) %>% 
  rename(from_state = t1_state) %>% 
  slice(match(order, from_state))

t1_counts <- as.data.frame(t1_counts)
rownames(t1_counts) <- t1_counts$from_state
t1_counts <- select(t1_counts,-c("from_state"))

summary_women <- t1_counts
summary_women

rm(t1_counts)
```

<span style="color:blue">Make prediction for $t_2$.</span>
```{r}
mmult <- as.matrix(trans_matrix_women) %*% as.matrix(summary_women)

summary_women <- summary_women %>%
  mutate(
    t2_pred = unlist(lapply(seq_len(ncol(mmult)), function(i) mmult[,i]))
  ) 
summary_women
```

<span style="color:blue">Count the actual cells for $t_2$.</span>
```{r}
t2_counts <- transitions %>% 
  filter(woman==1) %>%
  count(t2_state, name = "t2_count") %>% 
  filter(t2_state %in% c("Junior","Senior","Exit")) %>% 
  rename(from_state = t2_state) %>% 
  slice(match(order, from_state))

t2_counts <- as.data.frame(t2_counts)
rownames(t2_counts) <- t2_counts$from_state
t2_counts <- select(t2_counts,-c("from_state"))
t2_counts

summary_women <- cbind(summary_women, t2_counts)
summary_women

rm(t2_counts)
```

<span style="color:blue">Calculate the % difference.</span>
```{r}
summary_women <- summary_women %>%
  mutate(
    percent_diff = (t2_pred - t2_count)/t2_count
  ) 
summary_women
```


## Prediction for $t_2$ - men
<span style="color:blue">Zooming into women, let's count actual cells for $t_1$.</span>
```{r}
t1_counts <- transitions %>% 
  filter(woman==0) %>%
  count(t1_state, name = "t1_count") %>% 
  filter(t1_state %in% c("Junior","Senior","Exit")) %>% 
  rename(from_state = t1_state) %>% 
  slice(match(order, from_state))

t1_counts <- as.data.frame(t1_counts)
rownames(t1_counts) <- t1_counts$from_state
t1_counts <- select(t1_counts,-c("from_state"))

summary_men <- t1_counts
summary_men

rm(t1_counts)
```

<span style="color:blue">Make prediction for $t_2$.</span>
```{r}
mmult <- as.matrix(trans_matrix_men) %*% as.matrix(summary_men)

summary_men <- summary_men %>%
  mutate(
    t2_pred = unlist(lapply(seq_len(ncol(mmult)), function(i) mmult[,i]))
  ) 
summary_men
```

<span style="color:blue">Count the actual cells for $t_2$.</span>
```{r}
t2_counts <- transitions %>% 
  filter(woman==0) %>%
  count(t2_state, name = "t2_count") %>% 
  filter(t2_state %in% c("Junior","Senior","Exit")) %>% 
  rename(from_state = t2_state) %>% 
  slice(match(order, from_state))

t2_counts <- as.data.frame(t2_counts)
rownames(t2_counts) <- t2_counts$from_state
t2_counts <- select(t2_counts,-c("from_state"))
t2_counts

summary_men <- cbind(summary_men, t2_counts)
summary_men

rm(t2_counts)
```

<span style="color:blue">Calculate the % difference.</span>
```{r}
summary_men <- summary_men %>%
  mutate(
    percent_diff = (t2_pred - t2_count)/t2_count
  ) 
summary_men
```


## Regression to predict attrition
```{r}
# Collect Data for gender from Applications Dataset
Test_df = applications[ , c('patex_id', 'gender')]
#View(Test_df)

# Find only distinct values
Test_df = (distinct(Test_df))

# 5,649 values. Equal to the transitions dataset
# Merge dataset to transitions dataset

# merge two data frames by ID
New_df <- merge(transitions,Test_df,by="patex_id")
#View(New_df)

# Count null values
skim(New_df)
```
<span style="color:blue">so t0 is missing 1250 values, gender is missing 663, but exit data is missing the most with 4738 values.</span>

<span style="color:blue">Replace missing vales with random gender.</span>
```{r}
attach(New_df)
table(gender)
# 3438 men and 1548 women. So there are 0.69 propration of woman and 0.31 of men.
```

```{r}
# Replace all missing values at propotional rate
y = sum(is.na(gender))
New_df$gender[is.na(New_df$gender)] <- sample(c('male','female'),  prob=c(0.69, 0.31),y, replace = TRUE)

# Drop null values in t2
#New_df <- New_df[is.na(New_df$t2_state),] 
New_df = New_df[!(is.na(New_df$t2_state)| New_df$t2_state==""),]

# Drop New hires in t2
New_df<-New_df[New_df$t2_state!="New hire",]
```


<span style="color:blue">Creating an quadratic discriminat regression</span>
<span style="color:blue">We only have three values in t2_state. Junior, Senior and Exit.</span>
```{r}
library('fastDummies')
New_df <- dummy_cols(New_df, select_columns = 'gender')
attach(New_df)

library(MASS)
#myqda = qda(t2_state~gender_male+tenure_t2+tenure_t1+tenure_t0)
```

<span style="color:blue">Creating an quadratic discriminat regression.</span>
```{r}
# Creating a numerical variable for attrition 
New_df$t2_state[New_df$t2_state=="Exit"]=1
New_df$t2_state[New_df$t2_state=="Junior"]=0
New_df$t2_state[New_df$t2_state=="Senior"]=0


New_df <- transform(New_df,t2_state = as.factor(as.numeric(New_df$t2_state)))
attach(New_df)
table(New_df$t2_state)
```


```{r}
# First Dirty Model
search()
logit=glm(t2_state ~ gender_male+tenure_t2, family = "binomial",data = New_df) 
summary(logit)
```

```{r}
library(ggplot2)
plot=ggplot(New_df, aes(x=tenure,y=t2_state))+geom_point()
line=geom_smooth(method="glm", formula=t2_state~gender_male+tenure_t2, method.args=list(family=binomial))
plot+line
```

```{r}
# Predictions for male and female attrition
b0=coef(logit)[1]
b1=coef(logit)[2]
b2=coef(logit)[3]
```

Let's look into the attrition of junior and senior, as well as women and men

```{r}
sex = 0 # female
ten = 1 # One year
m=exp(b0+b1*sex+b2*ten)
prob = m/(1+m)
prob
```
<span style="color:blue">25.7% Chance this junior woman quits.</span>

```{r}
sex = 1 # male
ten = 1 # One year
m=exp(b0+b1*sex+b2*ten)
prob = m/(1+m)
prob
```
<span style="color:blue">27.2% Chance this junior man quits.</span>

```{r}
sex = 0 # Female
ten = 10 # Ten years
m=exp(b0+b1*sex+b2*ten)
prob = m/(1+m)
prob
```
<span style="color:blue">8.1% Chance this senior woman quits.</span>


```{r}
sex = 1 # Male
ten = 10 # Ten years
m=exp(b0+b1*sex+b2*ten)
prob = m/(1+m)
prob
```
<span style="color:blue">9.1% chance this senior man will quit.</span>

The above aligns with our transition matrices that junior have a higher attrition rate than senior and women have a lower attrition rate than men. 




<span style="color:blue">Appendix: more exploratory work</span>
```{r}
#library('fastDummies')
#New_df <- dummy_cols(New_df, select_columns ='t1_state')
#attach(New_df)
```

```{r}
# Second Dirty Model
#search()
#logit2=glm(t2_state ~ gender_male+tenure_t2+t1_state_Junior+t1_state_Senior, family = "binomial",data = New_df) 
#summary(logit2)
```

```{r}
# Predictions for male and female attrition
#b0=coef(logit2)[1]
#b1=coef(logit2)[2]
#b2=coef(logit2)[3]
#b3=coef(logit2)[4]
#b4=coef(logit2)[5]
```

```{r}
#sex = 1 # Male
#ten = 10 # Ten years
# Senior at t1
#m=exp(b0+b1*sex+b2*ten+b4*1)
#prob = m/(1+m)
#prob
```

```{r}
#sex = 1 # Male
#ten = 1 # One year
# Junior at t1
#m=exp(b0+b1*sex+b2*ten+b3*1)
#prob = m/(1+m)
#prob
```

```{r}
#sex = 0 # Female
#ten = 10 # Ten years
# Senior at t1
#m=exp(b0+b1*sex+b2*ten+b4*1)
#prob = m/(1+m)
#prob
```

```{r}
#sex = 0 # Female
#ten = 1 # One year
# Junior at t1
#m=exp(b0+b1*sex+b2*ten+b3*1)
#prob = m/(1+m)
#prob
```

