---
title: "USPTO Exercise"
author: "Chelsea"
date: "1/25/2022"
output: html_document
---

```{r}
library(tidyverse)
library(lubridate)
library(stringr)
library(skimr)
library(gender)
library(wru)
library(dplyr)
library(plyr)
```

app_data_sample.parquet contains subsample data on patent applications
examiner_aus.csv contains data on Art Units associated with each examiner at each point in time
examiner_gs.csv contains data on examiner GS pay grade changes
examiner_ids.csv is a crosswalk table to link to the applications data

```{r}
examiner_aus = read.csv("examiner_aus.csv") 
examiner_gs = read.csv("examiner_gs.csv") 
examiner_ids = read.csv("examiner_ids.csv")

library(arrow)
app_data_sample = read_parquet("app_data_sample.parquet", as_tibble = TRUE)

examiner_aus %>% tbl_vars()
examiner_gs %>% tbl_vars()
examiner_ids %>% tbl_vars()
app_data_sample %>% tbl_vars()
```

##### DATA CLEANING #####

```{r}
# Format dates
examiner_gs$start_date = as.Date(examiner_gs$start_date, tryFormats = c("%m/%d/%y", "%m/%d/%Y"))
examiner_gs$end_date = as.Date(examiner_gs$end_date, tryFormats = c("%m/%d/%y", "%m/%d/%Y"))
```

##### 1. FEATURE ENGINEERING - CREATING INDIVIDUAL-LEVEL VARIABLES #####
```{r}
# Split examiner's name into first name and last name
examiners <- examiner_gs %>% 
  separate(examiner_name, c("surname", "firstname"))
attach(examiners)
```

```{r}
# Add a column gender based on examiner's first name using package 'gender'
# Consolidate all first names into a directory
consolidated <- distinct(data.frame(all_names = as.character(as.vector(as.matrix(firstname)))))
# Get the gender predictions for each first name
genderpred <- gender(consolidated$all_names, method = "ssa")
# Perform a join to keep those names even the gender package cannot process
genderpred <- genderpred %>%
  right_join(consolidated, by = c("name" = "all_names"))
# Look up gender from directory
examiners <- left_join(examiners, genderpred, c("firstname" = "name"))
examiners <- select(examiners,-c("proportion_male","proportion_female","year_min","year_max"))
attach(examiners)
rm(consolidated)
rm(genderpred)
```

```{r}
# Add a column race based on examiner's surname using package 'wru' (but unsure what to use as the county/tract/block code)
examiners$county <- "061"
examiners <- predict_race(examiners, census.surname = TRUE, surname.only = TRUE)
examiners$race <- colnames(examiners[10:14])[max.col(examiners[10:14], ties.method = "first")]
examiners$race = substr(examiners$race,6,8)
examiners <- select(examiners,-c("pred.whi","pred.bla","pred.his","pred.asi","pred.oth"))
```

```{r}
# Add a column tenure based on examiner's start date and end date using package 'lubridate'
library(lubridate)
examiners <- examiners %>%
  mutate(tenure = case_when(
    is.na(end_date) ~ lubridate::time_length(difftime(as.Date("2016-12-31"), start_date), "years"),
    TRUE ~ lubridate::time_length(difftime(end_date, start_date), "years")
  ))
```

```{r}
## Generate profile of unique examiners
library(dplyr)
detach(package:plyr)
examiner_profile <- examiners %>%
  group_by(surname, firstname, old_pid, new_pid, gender, race) %>%
  summarise(min_grade = min(examiner_grade),
            max_grade = max(examiner_grade),
            start_date = min(start_date),
            end_date = max(end_date),
            tenure = mean(tenure))
rm(examiners)
```

```{r}
# Get patex_id for each examiner
library(plyr)
examiner_profile <- join(examiner_profile, examiner_ids, by=c("old_pid","new_pid"), type="left",match = "first")
examiner_profile <- select(examiner_profile,-c("examiner_name"))
```

##### 2. FEATURE ENGINEERING - CREATING PANEL DATASET OF QUARTERLY AGGREGATED PRODUCTION MEASURES FOR EACH EXAMINER #####
```{r}
## Generate list of applications
# Create a new column for Year & Quarter
library(lubridate)
examiner_profile$qtr <- lubridate::quarter(examiner_profile$start_date, with_year = T)
applications <- app_data_sample
applications$qtr <- lubridate::quarter(applications$filing_date, with_year = T)
```

```{r}
# Get number of applications
library(plyr)
temp <- ddply(applications, .(examiner_id), nrow)
names(temp) <- c("examiner_id","n_applications")
examiner_profile <- left_join(examiner_profile, temp, by=c("patex_id"="examiner_id"), match="first")
```

```{r}
# Get current art unit
temp <- examiner_aus[rev(order(examiner_aus$year,examiner_aus$month)),]
temp <- temp[match(unique(c(temp$old_pid,temp$examiner_art_unit)), c(temp$old_pid,temp$examiner_art_unit)),]
temp <- select(temp, c("old_pid","examiner_art_unit"))
names(temp) <- c("old_pid","examiner_au")
examiner_profile <- left_join(examiner_profile, temp, by="old_pid", match="first")
```

```{r}
## Generate list of art unit
# Get number of people in art unit
temp <- examiner_aus[rev(order(examiner_aus$year,examiner_aus$month)),]
temp <- temp[match(unique(c(temp$old_pid,temp$examiner_art_unit)), c(temp$old_pid,temp$examiner_art_unit)),]
art_unit <- ddply(temp, .(examiner_art_unit), nrow)
names(art_unit) <- c("examiner_au","n_examiners")
```

```{r}
# Get number of women in art unit
library(plyr)
temp <- ddply(filter(examiner_profile, gender == "female"), .(examiner_au), nrow)
names(temp) <- c("examiner_au","n_female")
art_unit <- join(art_unit, temp, by="examiner_au", type="left", match = "first")
```

```{r}
# Get number of examiners by race in art unit
library(plyr)
temp <- ddply(examiner_profile, .(examiner_au, race), nrow)
names(temp) <- c("examiner_au","race","count")
temp <- temp %>%
  mutate(whi = 0, bla=0, asi=0, oth=0)
for (i in 1:length(temp$examiner_au)){
  if (temp$race[i]=="whi"){
    temp$whi[i]=temp$count[i]
  } else if (temp$race[i]=="bla"){
    temp$bla[i]=temp$count[i]
  } else if (temp$race[i]=="asi"){
    temp$asi[i]=temp$count[i]
  } else if (temp$race[i]=="oth"){
    temp$oth[i]=temp$count[i]
  }
}
library(dplyr)
detach(package:plyr)
temp <- temp %>%
  group_by(examiner_au) %>%
  summarise(n_whi = max(whi),
            n_bla = max(bla),
            n_asi = max(asi),
            n_oth = max(oth))
library(plyr)
art_unit <- join(art_unit, temp, by="examiner_au", type="left",match = "first")
```


```{r}
# Get separation indicator (examiner left the USPTO)
# all end dates in examiner_gs / examiner_profile are NA, all examiners are active employees
```

```{r}
# Get AU move indicator (examiner changed art units)
temp <- unique(select(examiner_aus,-c("year","month")))
temp <- ddply(temp, .(old_pid), nrow)
names(temp) <- c("old_pid","n_au")
examiner_profile <- left_join(examiner_profile, temp, by="old_pid", match="first")

examiner_profile <- examiner_profile %>%
  mutate(move_au = ifelse(examiner_profile$n_au>1,TRUE,FALSE))
#examiner_profile <- select(examiner_profile,-c("n_au"))
```

```{r}
# Export results to csv
write.csv(applications,"output_applications.csv")
write.csv(examiner_profile,"output_examiner_profile.csv")
write.csv(art_unit,"output_art_unit.csv")
```

