---
title: "USPTO Exercise"
author: "Chelsea"
date: "1/25/2022"
output: html_document
---

##### 0. DATA PREPROCESSING #####

```{r}
library(tidyverse)
library(lubridate)
library(stringr)
library(skimr)
library(gender)
library(wru)
library(dplyr)
library(plyr)
```

app_data_sample.parquet contains subsample data on patent applications
examiner_aus.csv contains data on Art Units associated with each examiner at each point in time
examiner_gs.csv contains data on examiner GS pay grade changes
examiner_ids.csv is a crosswalk table to link to the applications data

```{r}
examiner_aus = read.csv("examiner_aus.csv") 
examiner_gs = read.csv("examiner_gs.csv") 
examiner_ids = read.csv("examiner_ids.csv")

library(arrow)
app_data_sample = read_parquet("app_data_sample.parquet", as_tibble = TRUE)

examiner_aus %>% tbl_vars()
examiner_gs %>% tbl_vars()
examiner_ids %>% tbl_vars()
app_data_sample %>% tbl_vars()
```

```{r}
# Format dates
examiner_gs$start_date = as.Date(examiner_gs$start_date, tryFormats = c("%m/%d/%y", "%m/%d/%Y"))
examiner_gs$end_date = as.Date(examiner_gs$end_date, tryFormats = c("%m/%d/%y", "%m/%d/%Y"))
```

##### 1. FEATURE ENGINEERING - CREATING INDIVIDUAL-LEVEL VARIABLES #####

```{r}
# Split examiner's name into first name and last name
examiners <- examiner_gs %>% 
  separate(examiner_name, c("surname", "firstname"))
attach(examiners)
```

```{r}
# Add a column gender based on examiner's first name using package 'gender'
# Consolidate all first names into a directory
consolidated <- distinct(data.frame(all_names = as.character(as.vector(as.matrix(firstname)))))
# Get the gender predictions for each first name
genderpred <- gender(consolidated$all_names, method = "ssa")
# Perform a join to keep those names even the gender package cannot process
genderpred <- genderpred %>%
  right_join(consolidated, by = c("name" = "all_names"))
# Look up gender from directory
examiners <- left_join(examiners, genderpred, c("firstname" = "name"))
examiners <- select(examiners,-c("proportion_male","proportion_female","year_min","year_max"))
attach(examiners)
rm(consolidated)
rm(genderpred)
```

```{r}
# Add a column race based on examiner's surname using package 'wru' (but unsure what to use as the county/tract/block code)
examiners$county <- "061"
examiners <- predict_race(examiners, census.surname = TRUE, surname.only = TRUE)
examiners$race <- colnames(examiners[10:14])[max.col(examiners[10:14], ties.method = "first")]
examiners$race = substr(examiners$race,6,8)
examiners <- select(examiners,-c("pred.whi","pred.bla","pred.his","pred.asi","pred.oth"))
```

```{r}
# Add a column tenure based on examiner's start date and end date using package 'lubridate'
library(lubridate)
examiners <- examiners %>%
  mutate(tenure = case_when(
    is.na(end_date) ~ lubridate::time_length(difftime(as.Date("2016-12-31"), start_date), "years"),
    TRUE ~ lubridate::time_length(difftime(end_date, start_date), "years")
  ))
```

```{r}
## Generate profile of unique examiners
library(dplyr)
detach(package:plyr)
examiner_profile <- examiners %>%
  group_by(surname, firstname, old_pid, new_pid, gender, race) %>%
  summarise(min_grade = min(examiner_grade),
            max_grade = max(examiner_grade),
            start_date = min(start_date),
            end_date = max(end_date),
            tenure = mean(tenure))
rm(examiners)
```

```{r}
# Get patex_id for each examiner
library(plyr)
examiner_profile <- join(examiner_profile, examiner_ids, by=c("old_pid","new_pid"), type="left",match = "first")
examiner_profile <- select(examiner_profile,-c("examiner_name"))
```

##### 2. FEATURE ENGINEERING - CREATING PANEL DATASET OF QUARTERLY AGGREGATED PRODUCTION MEASURES FOR EACH EXAMINER #####

```{r}
## Generate list of applications
# Create a new column for Year & Quarter
library(lubridate)
examiner_profile$qtr <- lubridate::quarter(examiner_profile$start_date, with_year = T)
applications <- app_data_sample
applications$qtr <- lubridate::quarter(applications$filing_date, with_year = T)
```

```{r}
# Get number of applications
library(plyr)
temp <- ddply(applications, .(examiner_id), nrow)
names(temp) <- c("examiner_id","n_applications")
examiner_profile <- left_join(examiner_profile, temp, by=c("patex_id"="examiner_id"), match="first")
```

```{r}
library(plyr)
temp <- ddply(filter(applications, !startsWith(application_number,"90"), !startsWith(application_number,"95"), !startsWith(application_number,"96"))
              , .(examiner_id), nrow)
names(temp) <- c("examiner_id","n_new_applications")
examiner_profile <- left_join(examiner_profile, temp, by=c("patex_id"="examiner_id"), match="first")
```


```{r}
# Get number of abandoned applications
library(plyr)
temp <- ddply(filter(applications, disposal_type == "ABN"), .(examiner_id), nrow)
names(temp) <- c("examiner_id","n_ABN_applications")
examiner_profile <- left_join(examiner_profile, temp, by=c("patex_id"="examiner_id"), match="first")
```

```{r}
# Get number of allowed applications
library(plyr)
temp <- ddply(filter(applications, disposal_type == "ISS"), .(examiner_id), nrow)
names(temp) <- c("examiner_id","n_ISS_applications")
examiner_profile <- left_join(examiner_profile, temp, by=c("patex_id"="examiner_id"), match="first")
```

```{r}
library(plyr)
temp <- ddply(filter(applications, disposal_type == "PEND"), .(examiner_id), nrow)
names(temp) <- c("examiner_id","n_PEND_applications")
examiner_profile <- left_join(examiner_profile, temp, by=c("patex_id"="examiner_id"), match="first")
```


```{r}
# Get current art unit
temp <- examiner_aus[rev(order(examiner_aus$year,examiner_aus$month)),]
temp <- temp[match(unique(c(temp$old_pid,temp$examiner_art_unit)), c(temp$old_pid,temp$examiner_art_unit)),]
temp <- select(temp, c("old_pid","examiner_art_unit"))
names(temp) <- c("old_pid","examiner_au")
examiner_profile <- left_join(examiner_profile, temp, by="old_pid", match="first")
```

```{r}
## Generate list of art unit
# Get number of people in art unit
temp <- examiner_aus[rev(order(examiner_aus$year,examiner_aus$month)),]
temp <- temp[match(unique(c(temp$old_pid,temp$examiner_art_unit)), c(temp$old_pid,temp$examiner_art_unit)),]
art_unit <- ddply(temp, .(examiner_art_unit), nrow)
names(art_unit) <- c("examiner_au","n_examiners")
```

```{r}
# Get number of women in art unit
library(plyr)
temp <- ddply(filter(examiner_profile, gender == "female"), .(examiner_au), nrow)
names(temp) <- c("examiner_au","n_female")
art_unit <- join(art_unit, temp, by="examiner_au", type="left", match = "first")
```

```{r}
# Get number of examiners by race in art unit
library(plyr)
temp <- ddply(examiner_profile, .(examiner_au, race), nrow)
names(temp) <- c("examiner_au","race","count")
temp <- temp %>%
  mutate(whi = 0, bla=0, asi=0, oth=0)
for (i in 1:length(temp$examiner_au)){
  if (temp$race[i]=="whi"){
    temp$whi[i]=temp$count[i]
  } else if (temp$race[i]=="bla"){
    temp$bla[i]=temp$count[i]
  } else if (temp$race[i]=="asi"){
    temp$asi[i]=temp$count[i]
  } else if (temp$race[i]=="oth"){
    temp$oth[i]=temp$count[i]
  }
}
library(dplyr)
detach(package:plyr)
temp <- temp %>%
  group_by(examiner_au) %>%
  summarise(n_whi = max(whi),
            n_bla = max(bla),
            n_asi = max(asi),
            n_oth = max(oth))
library(plyr)
art_unit <- join(art_unit, temp, by="examiner_au", type="left",match = "first")
```


```{r}
# Get separation indicator (examiner left the USPTO)
# all end dates in examiner_gs / examiner_profile are NA, all examiners are active employees
```

```{r}
# Get AU move indicator (examiner changed art units)
temp <- unique(select(examiner_aus,-c("year","month")))
temp <- ddply(temp, .(old_pid), nrow)
names(temp) <- c("old_pid","n_au")
examiner_profile <- left_join(examiner_profile, temp, by="old_pid", match="first")

examiner_profile <- examiner_profile %>%
  mutate(move_au = ifelse(examiner_profile$n_au>1,TRUE,FALSE))
#examiner_profile <- select(examiner_profile,-c("n_au"))
```

```{r}
# Export results to csv
write.csv(applications,"output_applications.csv")
write.csv(examiner_profile,"output_examiner_profile.csv")
write.csv(art_unit,"output_art_unit.csv")
```

##### 4. EXPLORING REGRESSION #####

```{r}
### Count null values
df = examiner_profile
summary(df)
table(is.na(df))
attach(df)
```

Many null values due to type of dataset (2.6%)
Can't really drop them so I will replace them with mean to make

```{r}
### Find all columns with missing values
test_df = df[,c("n_applications","n_new_applications", 
"n_ABN_applications", "n_ISS_applications",
"n_PEND_applications")]

list_na <- colnames(test_df)[ apply(test_df, 2, anyNA) ]
```

```{r}
### Find column means
average_missing <- apply(test_df[,colnames(test_df) %in% list_na],
                         2,
                         mean,
                         na.rm =  TRUE)
average_missing
```

```{r}
# Replace missing values with mean value
library(dplyr)
df_replace <- df %>%
  mutate(replace_mean_na  = ifelse(is.na(n_applications), average_missing[1], n_applications),
         replace_mean_nn = ifelse(is.na(n_new_applications), average_missing[2], n_new_applications),
         replace_mean_n_AB = ifelse(is.na(n_ABN_applications), average_missing[3], n_ABN_applications),
         replace_mean_n_ISS = ifelse(is.na(n_ISS_applications), average_missing[4], n_ISS_applications),
         replace_mean_n_PEND = ifelse(is.na(n_PEND_applications), average_missing[5], n_PEND_applications))

sum(is.na(df_replace$replace_mean_na))
sum(is.na(df_replace$replace_mean_nn))
sum(is.na(df_replace$replace_mean_n_AB))
sum(is.na(df_replace$replace_mean_n_ISS))
sum(is.na(df_replace$replace_mean_n_PEND))

View(df_replace)
```

```{r}
# Dummify categorical variables
library(fastDummies)
df_replace = dummy_cols(df_replace,select_columns = 'race')
df_replace = dummy_cols(df_replace,select_columns = 'gender')
```

Now that all variables are cleaned and ready
We can now start testing
Before we get to that, lets adaboost

```{r}
library(randomForest)
library(gbm) 
set.seed (1)
attach(df_replace)
boosted=gbm(tenure~min_grade+max_grade+qtr+replace_mean_na+replace_mean_nn+replace_mean_n_AB
            +replace_mean_n_ISS+replace_mean_n_PEND+race_asi+race_bla+race_his
            +race_oth+race_whi+gender_male+gender_female,data=df_replace,distribution= "gaussian",n.trees=10000, interaction.depth=4)
summary(boosted)
```

This model displayed that qtr is highly influential so I am cutting it

```{r}
boosted=gbm(tenure~min_grade+max_grade+replace_mean_na+replace_mean_nn+replace_mean_n_AB
            +replace_mean_n_ISS+replace_mean_n_PEND+race_asi+race_bla+race_his
            +race_oth+race_whi+gender_male+gender_female,data=df_replace,distribution= "gaussian",n.trees=10000, interaction.depth=4)
summary(boosted)
```

This model indicated that the only important variables were min_grade and max_grade so I am cutting those as well

```{r}
boosted=gbm(tenure~replace_mean_na+replace_mean_nn+replace_mean_n_AB
            +replace_mean_n_ISS+replace_mean_n_PEND+race_asi+race_bla+race_his
            +race_oth+race_whi+gender_male+gender_female,data=df_replace,distribution= "gaussian",n.trees=10000, interaction.depth=4)
summary(boosted)
```

All values replaced with their respective means had significant influence
Now on to building Linear model

```{r}
lr.fit= lm(tenure~max_grade+min_grade+gender_male+gender_female+race_whi+replace_mean_na+replace_mean_n_AB
           +replace_mean_n_ISS+replace_mean_n_PEND,data = df_replace)
summary(lr.fit)
```

```{r}
sm<-summary(lr.fit)
mse <- function(sm) 
mean(sm$residuals^2)
```
MSE=28.45
This model is terrible but improves if we insert min_grade and max_grade (from MSE = 50.92 to 28.45)

```{r}
### Plotting some of the data
library(car)
residualPlots(lr.fit)

qqPlot(lr.fit)
```

To estimate turnover, I borrowed an equation from the American Staffing Association
If we table quarters, we can estimate how many employees were at the company for each
quarter. If we sum this by year and then we divide this number by 4 for number of quarters in a year
we get average employment. If we sum tenure for all employees of that year and then divide
that by average employment, we can develop an estimate for turnover

```{r}
### Calculate total amount by quarter
df_replace$qtr <- as.character(df_replace$qtr)

df_replace$count <- as.numeric(ave(df_replace$qtr, df_replace$qtr, FUN = length))
df_replace
```

```{r}
### Calculate total tenure by quarter
ag = aggregate(df_replace$tenure, by=list(df_replace$qtr), FUN=sum)

df_replace %>% group_by(qtr) %>% summarise(max_val = max(tenure),
                                    max_date = qtr[which.max(tenure)])
```

